---
title: "lab_08"
author: "derek willis"
date: "2023-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. We'll be making some charts, working with dates and retrieving Census data.
```{r}
library(tidyverse)
library(lubridate)
library(ggthemes)
library(tidycensus) 
```


## Load and modify data

**Task** Create a codeblock and load the following data from the data folder:

* Prince George's County 911 Overdose Calls
```{r}
pg_county911calls <- read_csv("data/prince_georges_2022_overdoses.csv")

pg_county911calls <- pg_county911calls |>
  mutate(month = month(datetime, label=TRUE)) |>
  mutate(
    date = date(datetime),
    week = week(datetime),
    month = month (datetime, label=TRUE)
  )
    


 
```

You will need to create columns for the date, week and month based on the existing `datetime` column.



## Questions 

**Q1.** Which month saw the greatest percentage of total calls? Create a dataframe that calculates the percentage of all calls that each month's calls represents. Make a bar chart from that dataframe. Your bar chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable bars - the values shouldn't be overlapping

Compare the results here to those from the pre_lab_08 bar chart - are there major differences in the months with the highest and lowest figures? Describe that below.

**A1.** 
In the bar graph that we used for the Baltimore County 911 calls in the pre-lab, December was the second lowest month for OD calls, whereas in PG County, it is the month with the most amount of calls. Also, August was the month where Baltimore County saw the most amount of 911 OD calls, and in this dataset it is the second lowest. 

In general, Baltimore County saw more OD calls during the Summer months and PG saw most of their overdose calls towards the latter part of the year (Dec, Nov, Sept), with the escpetion of July.
```{r}
pg_county911calls_month <- pg_county911calls |>
  group_by(month) |>
  summarise(total_calls = n())

pg_county911calls_month |>
 mutate(rate_of_calls = (total_calls/1397)*100) |>
   arrange(desc(total_calls))

pg_county911calls_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month, total_calls), weight=total_calls)) +
  theme_economist() +
  labs(
    title="PG County 911 Overdose Calls by month",
    x = "month",
    y = "total calls",
    caption = "source: PG County EMS" 
    
  )

  
 
```

**Q2.** Let's visualize this data on a weekly basis using a line chart. As in Q1, generate a dataframe with the total number of calls for each week, and then create a line chart to show the distribution of calls over time. Your line chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable labels

Describe the pattern of results; you may want to look at the data to dig into particular outliers.

**A2.** 
You can see OD calls reached its peak in the middle of the year, which is during the summer. Then as we got towards the end of the year, you can clearly see and stop drop off in 911 OD calls. However, it is interesting to see that the right after the spike in the middle of the year, the number of calls dropped to one of the lowest all year.Additionally, right before the steep dropoff, the number of OD calls reached its second highest total all year.

```{r}
pg_county911calls_week <- pg_county911calls |>
  group_by(week) |>
  summarise(total_calls = n())

pg_county911calls_week |>
  ggplot() + 
  geom_line(aes(x=week, y=total_calls)) +
  labs(
    title="PG County 911 Overdose Calls by week",
    x = "week",
    y = "total calls",
    caption = "source: PG County EMS" 
    
  )
```

**Q3.**  A scatterplot is a type of chart that helps us see relationships between two variables. One variable goes on the x axis, the other on the y axis.  For each row/observation in our data, a scatterplot puts a circle (or a "point") where the two variables intersect on a grid. 

Statisticians use scatterplots to show graphically whether one variable is correlated -- related, in a statistical sense -- with another variable.  A classic example is the [relationship between ice cream sales and temperature](https://www.mathsisfun.com/data/scatter-xy-plots.html). The scatterplot below -- press play to load the image -- shows that relationship, that an increase in temperature is associated with an increase in ice cream sales. When it's 12C, sales are 200 dollars, and when it's hotter, 25C, sales are 600 dollars.

```{r}
knitr::include_graphics("https://www.mathsisfun.com/data/images/scatter-ice-cream1.svg")
```

We're going to use a scatterplot a little differently, to get a visual sense of two key variables: 

Our question is: does the median income in a zip code have any relationship to the number of overdose 911 calls in that zip code?

To answer this question, do the following:

1. Generate a dataframe with the number of 911 calls for each zip code.
2. Get data from the Census Bureau showing median household income for Maryland zip codes.
3. Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.
4. Make a scatterplot showing the total calls and median income. I didn't show you how to do this, so look it up! Googling "ggplot scatterplot" is a good start.
5. Give it an appropriate title, source, and x and y axis titles.
6. Add a label for each point that shows the zip code using geom_text() - see some examples of its use at https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples. Try to make the names as easy to read as possible by avoiding overlap.
7. In the answer space below, describe what you see and answer the questions posed above. In a general sense, what do you think this means? Feel free to consider the actual raw values: how would you report out the main point(s) of this chart?

**A3.**  
*I asked ChatGPT 'how do I create a scatterplot using tidyverse showing total calls and income' and then 'how do I then switch the dots on a scatterplot to text values from a column in a dataset in r studio with tidyverse' and then my groupmates Cait and April assisted me with some of the error messages that I was getting.*

I would say that for the most part, the higher the median income, the lower amount of overdose calls that you typically see from that zip code. However, one of the counties with the lowest amount of calls, and the county with the highest amount of calls have almost the same median income, with very similar demographics. So I wonder what the correlation is there. 
```{r}
pg_county911calls_withzip <- pg_county911calls |>
  group_by(zipcode) |>
  summarise(total_calls = n())
  
md <- get_acs(geography = "zcta",
              variables = c(medincome = "B19013_001"),
              state = "MD",
              year = 2019)

pg_county911calls_withincome <- pg_county911calls_withzip |> 
  left_join(md, by=c('zipcode'='GEOID'))

ggplot(data = pg_county911calls_withincome, aes(x = total_calls, y = estimate )) + 
  geom_text(aes(label = zipcode), size = 3, vjust = 0, check_overlap = TRUE) +
  geom_point(shape = 19, color = "purple", size = 0.5) +
    labs(title = "Total amount of 911 OD Calls vs. Median Income",
    x= "Total Calls", 
    y= "Income",
caption = "PG County EMS & 2019 Census"
) 




```
