---
title: "lab_03"
author: "Derek Willis"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(janitor)
# Load janitor
library(tidyverse)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the column NAMES so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=373564) |>
  clean_names()
# Load in the Maryland grants and loan data
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? Look at & search the records in the dataframe while you construct your code.

A1:
32 records
```{r}
pg_schools <- payments |>
  filter(
    str_detect(vendor_name, "PRINCE GEORGES PUBLIC SCHOOLS") |
    str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL") |
    str_detect(vendor_name, "BD OF EDUCATION OF PG CO")
    
  )
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: 2020 had the largest total amount, I would say I am surprised by the fact that there wasn't a higher jump between 2011 and 2014.

```{r}
pg_schools |>
  group_by(fiscal_year) |>
  summarize(total_amount = sum(Amount)) |>
  
  arrange(desc(total_amount))


```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3.

It would need to be cleaned up to not show duplicates of the same company. For example, we see Whiting Turner Contracting Co. and Whiting Turner Contracting Company. Also, A & I INC., A&I INC and A&I INC. Lastly, A Morton Thomas & Assoc INC and A Morton Thomas Associates INC.

To fix this, we could use the dup function and then maybe combine all by the same 5 digit zip code. 

```{r}
payments |>
  filter(agency_name == 'CAPITAL PROJECTS- UNIVERSITY OF MARYLAND') |>
  group_by(vendor_name) |>
  summarize(total_amount = sum(amount)) |>
  
  arrange(vendor_name)
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4:
Honestly, it looks like the grant funding has been sporadic. When looking at the grants from greatest to least greatest, the amount given compared to the year doesn't really have a trend from what I can see. 

```{r}
cleaned_md_grants_loans |> 
  filter(description == "STEM CELL RESEARCH GRANT") |>
  group_by(fiscal_year) |>
  
  arrange(desc(amount))

```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: There are 38 grantees in the zip code of 20740 in total. Again, we have a case where it seems like there are multiple or duplicate grantees. For example, there are three different columns where it says the source is DEPARTMENT OF BUDGET AND MANAGEMENT/DGS CAPITAL BUDGET. However, this could be a case where the same company made multiple donations, or there were split into multiple different payments. The most interesting one to me is DEPARTMENT OF BUSINESS AND ECONOMIC DEVELOPMENT/MD STATE ARTS COUNCIL b/c the grantee is listed as Elizabeth Gallauresi, and that seems to be the only grantee that is not a business but an individual. 

Additionally, when I searched up college park in the cleaned_md_grants_loans data, I did see three results that was included in the College Park description but had the zip code of 20742. I did orginally attempt to add that to the code as well, but for some reason I kept getting an error message so I discarded it. 

```{r}
cleaned_md_grants_loans |>
  filter(zip_code == 20740) |>
  group_by(source)
  
```

